Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2015-11-10 WSSSPE3/report/wg_main_funding_spec_expert.tex

One strategy to promote stability is implementing a mechansim for developers to obtain academic
credit for software development work. With such a mechanism in place, traditional academic funding
models and career tracks could properly sustain individuals for whom software development is their
primary contribution to research. A contributing factor to the problem with the current academic reward system is the
devastating effect on an academic
publication record resulting from time in industry; such postings often develop exactly the skills that research programmers
need, yet returns to university positions following an industry role are penalized by the current structures. Retention of
senior programmers is hard, because these people are highly in demand by the economy. However, people who have a PhD in
science and enter industry, may desire to return for diverse reasons, and should be welcomed back.

While development of new mechanisms in the current academic reward system is a worthy aspirational goal, such a dramatic
change in this structure does not seem likely in a time scale relevant to this working group. Accordingly, our working party
sought alternate solutions that may be achievable within the context of existing academic structures. The group felt that
developing dedicated research programming roles within the University, and finding stable funding for those individuals is
the most promising mechanism for creating a stable software development staff.

Measures of impact and success for research programming groups, as well as for individual research software engineers, will
be required in order to make the case to the University for continued funding. Research programmers will not be measured by publications, we hope, but by other measures. (what measures?) Middle-author publications are common for RSEs. Most RSEs welcome co-authorship on papers where the PIs consider the contribution deserves it.
(these last two statements seem contradictory)

@@

It is hard for an individual PI in a university or college to support dedicated research programming resources, as the need
for, and funding for, these activities is intermittent within the research cycle. To sustain this capacity, therefore, it is
necessary to aggregate this work across multiple research groups.

One solution is to fund dedicated programming roles for major research software projects at national laboratories
or other non-educational institutions. This solution is in place and working well for many well-used scientific codebases.
However, this strategy has limited application, as much of the body of software is created and maintained in research
universities. Therefore, we argue that research institutions should develop hybrid academic-technical tracks for this
capacity, where employees in this track work with more than one PI, rather than the traditional RA role within a single group
. This could be coordinated centrally, as a core facility, perhaps within research computing organizations which have
traditionally supported university cyberinfrastructure, libraryorganizations, or research offices. Alternatively, these
groups could be organizationally closer to research groups, sitting within academic departments. The most effective model
will vary from institution to institution, but the mandate and ways of working should be similar.

Having convinced ourselves that this would be a positive innovation, we were then faced with the specific question of how to
fund the initiation of this activity. A self-sustaining research software group will support itself through collaborations
with PIs in the normal grant process, with PIs choosing to fund some amount of research programming effort through grants in
the usual way. However, to bootstrap such a function to a level where it has sufficent reputation and client base to be self
-sustaining will generally require seed investment.

@@

We finally noted that the institutional case for such groups is made easier by having successful examples to point to. In the
UK, a collective effort to identify this community, with individuals clearly stating ``I'm a research software engineer'',
has been important to the campaign. It will be useful to the global effort to similarly identify emerging research software
organizations, and also, importantly, to identify longer-running research software groups, which have in some cases had a
long running \emph{sui-generis} existence, but which now can be identified as part of a wider solution. There remains the problem of how to "sell" the value of this investment to investigators in the university will be problem left to the individual universities, once the initial investment is made. This is an issue best addressed by th eindividual organizations that embark on the plan.

>>>>

One strategy to promote stability is implementing a mechansim for developers to obtain academic
credit for software development work. With such a mechanism in place, traditional academic funding
models and career tracks could properly sustain individuals for whom software development is their
primary contribution to research. A contributing factor to the problem with the current academic reward system is the
devastating effect on an academic
publication record resulting from time in industry; such postings often develop exactly the skills that research software
engineers need, yet returns to university positions following an industry role are penalized by the current structures.
Retention of senior developers is hard, because these people are highly in demand by the economy. However, people who have a
PhD in science and enter industry, may desire to return for diverse reasons, and should be welcomed back.

While development of new mechanisms in the current academic reward system is a worthy aspirational goal, such a dramatic
change in this structure does not seem likely in a time scale relevant to this working group. Accordingly, our working party
sought alternate solutions that may be achievable within the context of existing academic structures. The group felt that
developing dedicated research software engineering roles within the University, and finding stable funding for those individuals is the most promising mechanism for creating a stable software development staff.

Measures of impact and success for research programming groups, as well as for individual research software engineers, will
be required in order to make the case to the University for continued funding. Research software engineers will not be measured by publications, we hope, but by other measures. (what measures?) Middle-author publications are common for RSEs. Most RSEs welcome co-authorship on papers where the PIs consider the contribution deserves it.
(these last two statements seem contradictory, not sure which message is to be used?)

@@

It is hard for an individual PI in a university or college to support dedicated research software engineering resources, as
the need for, and funding for, these activities is intermittent within the research cycle. To sustain this capacity, therefore, it is necessary to aggregate this work across multiple research groups.

One solution is to fund dedicated software engineering roles for major research software projects at national laboratories
or other non-educational institutions. This solution is in place and working well for many well-used scientific codebases.
However, this strategy has limited application, as much of the body of software is created and maintained in research
universities. Therefore, we argue that research institutions should develop hybrid academic-technical tracks for this
capacity, where employees in this track work with more than one PI, rather than the traditional RA role within a single group
. This could be coordinated centrally, as a core facility, perhaps within research computing organizations which have
traditionally supported university cyberinfrastructure, libraryorganizations, or research offices. Alternatively, these
groups could be organizationally closer to research groups, sitting within academic departments. The most effective model
will vary from institution to institution, but the mandate and ways of working should be similar.

Having convinced ourselves that this would be a positive innovation, we were then faced with the specific question of how to
fund the initiation of this activity. A self-sustaining research software group will support itself through collaborations
with PIs in the normal grant process, with PIs choosing to fund some amount of research software engineering effort through grants in
the usual way. However, to bootstrap such a function to a level where it has sufficent reputation and client base to be self
-sustaining will generally require seed investment.

@@

We finally noted that the institutional case for such groups is made easier by having successful examples to point to. In the
UK, a collective effort to identify the research software engineering community, with individuals clearly stating "I'm a research software engineer", has been important to the campaign. It will be useful to the global effort to similarly identify emerging research software organizations, and also, importantly, to identify longer-running research software groups, which have in some cases had a long running \emph{sui-generis} existence, but which now can be identified as part of a wider solution. There remains the problem of how to "sell" the value of this investment to investigators within the university. This is an issue best addressed by the individual organizations that embark on the plan.
